Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-28 Paper_Draft2/Draft2.Rmd
trying to fix it
reverting to commit before latex issues and forcing push
git force

**Statistical Parity Difference** - This metric measures the difference between privileged and marginalized groups’ likelihood to get a particular outcome. The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

@@

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between 0.8 to 1.25 is considered fair [@Ronaghan2019AI].

@@

**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group [@GoogleDev].

@@

**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants [@aif360-oct-2018].

>>>>

## Statistical Parity Difference
This metric measures the difference between privileged and marginalized groups’ likelihood to get a particular outcome. The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

@@

## Disparate Impact Ratio
This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between 0.8 to 1.25 is considered fair [@Ronaghan2019AI].

@@

## Equal Opportunity Difference
This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group [@GoogleDev].

@@

## Average Odds Difference
This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants [@aif360-oct-2018].

### 2022-04-28 final_metrics_analysis.Rmd

# for Amrita
#use_condaenv("anaconda3")
```

@@

## Reweighting

@@

@@

>>>>

# for Amrita
# use_condaenv("anaconda3")
```

@@

Pre-processing refers to mitigating bias within the training data. For our pre-processing approach we are using AIF360's Reweighing algorithm. The reweighing approach "assigns different weights to the examples based upon their categories of protected attribute and outcome such that bias is removed from the training dataset" ( [Mahmoudian](https://towardsdatascience.com/reweighing-the-adult-dataset-to-make-it-discrimination-free-44668c9379e8)).

## Reweighing

@@

In-processing mitigates bias in classifiers while building a model. The in-processing technique we use is the prejudice remover algorithm from AIF360. The prejudice remover is a method for reducing indirect prejudice (i.e. how the COMPAS algorithm is racially biased because it uses proxy variables for race). This approach implements two different regularizers, one to avoid overfitting and one to enforce fair classification ( [Kamishima et. al.](https://link.springer.com/content/pdf/10.1007/978-3-642-33486-3_3.pdf)). The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by [Kamishima et. al](https://link.springer.com/content/pdf/10.1007/978-3-642-33486-3_3.pdf). This in turn enforces a classifier’s independence from sensitive information (like race).

@@

Our last approach, post-processing bias mitigation, is used after training a model. Post-processing algorithms equalize the outcomes (i.e., predicted classification values) to mitigate bias instead of adjusting the classifier or the training data( [Baxter](https://www.infoworld.com/article/3630450/what-is-ai-bias-mitigation-and-how-can-it-improve-ai-fairness.html)). We use calibrated equalized odds, which "optimizes over calibrated classifier score outputs to find probabilities with which to change output labels with an equalized odds objective" ( [AIF360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.algorithms.postprocessing.CalibratedEqOddsPostprocessing.html#aif360.algorithms.postprocessing.CalibratedEqOddsPostprocessing)). An equalized odds objective constrains classification algorithms such that no error type (false-positive or false-negative) disproportionately affects any population subgroup; both groups, in our case both white and Black defendants, should have the same false-positive and false-negative rates. Through the calibrated equalized odds method, we want to decrease bias while also maintaining calibration ( [Pleiss et. al.](https://arxiv.org/pdf/1709.02012.pdf)). [Calibration](https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555) refers to improving a model so that the distribution of predicted outcomes is similar to the distribution of observed probability in the training data.

## Calibrated Equalized Odds
